---
title: "Reproducible Science & Figures Assignment"
output:
  html_document: default
  pdf_document: default
date: "2023-12-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## QUESTION 01: Data Visualisation for Science Communication

### a) Provide your figure here:

```{r bad figure code, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
options(repos = c(CRAN = "https://cran.r-project.org/"))

# Install packages:
install.packages(c("ggplot2", "palmerpenguins", "janitor", "dplyr"))

# Load the packages:
library(ggplot2)
library(palmerpenguins)
library(janitor)
library(dplyr)

# Look at the table with the raw data
head(penguins_raw)

# Export data from the palmerpenguins package
write.csv(penguins_raw, "data/penguins_raw.csv")

# Load data from working directory
penguins_raw <- read.csv("data/penguins_raw.csv")

# Define Function for cleaning
clean_function <- function(penguins_data) {
    penguins_data %>%
        select(-starts_with("Delta")) %>%
        select(-Comments) %>%
        clean_names()
}

# Call Function
penguins_clean <- clean_function(penguins_raw)

# Check the column names in the new data frame
names(penguins_clean)

# filter the data
body_mass_clean <- penguins_clean %>%
  select(body_mass_g, flipper_length_mm, species, sex) %>%
  na.omit()

# Create the bad graph 
bodymass_scatterplot <- ggplot(data = body_mass_clean, aes(x = flipper_length_mm, y = body_mass_g)) + 
  geom_point(size = 3, aes(fill = sex)) +
    ylab("body mass") +
    xlab("flipper length") +
    ylim(c(0, 7000)) +
    xlim(c(0, 240))

print(bodymass_scatterplot)

# There is no need to provide the code for your bad figure, just use echo=FALSE so the code is hidden. Make sure your figure is visible after you knit it. 

```
```{r, echo=FALSE, message=FALSE}
# Create the bad graph 
bodymass_scatterplot <- ggplot(data = body_mass_clean, aes(x = flipper_length_mm, y = body_mass_g)) + 
  geom_point(size = 3, aes(fill = sex)) +
    ylab("body mass") +
    xlab("flipper length") +
    ylim(c(0, 7000)) +
    xlim(c(0, 240))

print(bodymass_scatterplot)

# There is no need to provide the code for your bad figure, just use echo=FALSE so the code is hidden. Make sure your figure is visible after you knit it. 
```



### b) Write about how your design choices mislead the reader about the underlying data (200-300 words).

The intention of this figure is to illustrate the relationship between body mass and flipper size across three Brush-tailed penguin species. However, there are several issues with the design of the plot that make the relationship difficult to interpret. The graph lacks a title and so the reader cannot discern whether it represents a correlation analysis or a linear regression analysis. A correlation analysis measures the strength and direction of a linear relationship, whereas a linear regression analysis measures the effect of an independent variable upon a dependent variable, and so the two must be distinguished [1]. The current arrangement of axes may suggest to the reader that flipper size is the explanatory variable, however, this is not necessarily the case. The labels lack units and therefore the numerical scales for both variables are arbitrary, preventing the reader from conceptualizing the data. The scales themselves extend beyond what is needed to fully encapsulate the distributions of both variables, and hence the results are densely clustered. This is spatially inefficient, and does not illustrate the spread of the data to the reader, which in turn makes the nature of the relationship more difficult to interpret. The formatting of data points is problematic, as their size is too large to distinguish between individual points, which again prevents the reader from observing the spread of data [2]. As well as this, there is no colour differentiation between male and female data points, which renders the legend useless, and prevents the reader from observing any sexual dimorphisms in these two physiological traits. There is also no way of deciphering what species a data point belongs to, and this omits potentially important interspecific variation. Finally, a trend line could have been added to suggest a pattern in the data to the reader before beginning hypothesis testing.


References:

[1] Bewick, V., Cheek, L. and Ball, J., 2003. Statistics review 7: Correlation and regression. Critical care, 7, pp.1-9.

[2] Micallef, L., Palmas, G., Oulasvirta, A. and Weinkauf, T., 2017. Towards perceptual optimization of the visual design of scatterplots. IEEE transactions on visualization and computer graphics, 23(6), pp.1588-1599.

------------------------------------------------------------------------

## QUESTION 2: Data Pipeline

------------------------------------------------------------------------

### Introduction

The palmerpenguins dataset contains size measurements for three penguin species observed on three islands in the Palmer Archipelago, Antartica. Gentoo, Adelie, and Chinstrap penguins are the three extant members of the Pygoscelis genus, also known as Brush-tail penguins. Mitochondrial and nuclear DNA evidence suggest that the genus split from other penguins around 38 million years ago. Gentoo and Chinstrap penguins comprise a sister group, as Adelie penguins diverged from other members of the genus around 19 million years ago [3].

Penguins are flightless birds, with wings morphed by evolution into flippers, used to propel themselves underwater. It is reasonable to assume that larger penguins require larger flippers in order to create more powerful swimming strokes. This study aims to evidence this proposed association between body mass and flipper length.

Adelie penguins are a sexually dimorphic species, with adult males averaging larger than adult females in mass (~11%) and flipper length (~3%) [4]. As the male and female phenotype differs, it is possible that the strength of relationship between body mass and flipper length also differs between the sexes. An additional aim of this study is to test for a difference in correlation of these two variables, between male and female penguins.


References:

[3] Baker, A.J., Pereira, S.L., Haddrath, O.P. and Edge, K.A., 2006. Multiple gene evidence for expansion of extant penguins out of Antarctica due to global cooling. Proceedings of the Royal Society B: Biological Sciences, 273(1582), pp.11-17.

[4] Jennings, S., Varsani, A., Dugger, K.M., Ballard, G. and Ainley, D.G., 2016. Sex-based differences in Adélie penguin (Pygoscelis adeliae) chick growth rates and diet. PLoS One, 11(3), p.e0149090.



```{r, echo=TRUE, warning=FALSE, message=FALSE}

# Install packages
# install.packages(c("ggplot2", "palmerpenguins", "janitor", "dplyr", "ragg"))

# Load packages
library(ggplot2)
library(palmerpenguins)
library(janitor)
library(dplyr)
library(ragg)

# Load functions into rmd file
source("functions/cleaning.r")

# Export data from the palmerpenguins package
write.csv(penguins_raw, "data/penguins_raw.csv")

# Load data from working directory
penguins_raw <- read.csv("data/penguins_raw.csv")

# View raw data
names(penguins_raw)

# Clean data using pre-loaded cleaning functions
penguins_clean <- penguins_raw %>% 
  clean_column_names() %>%     # simplifies column names
  shorten_species() %>%        # simplifies species names
  remove_empty_columns_rows()  # Removes any empty columns or rows

# View cleaned data
names(penguins_clean)

# Save cleaned data to working directory
write.csv(penguins_clean, "data/penguins_clean.csv")

# Filter data using pre-loaded sub-setting functions
bodymass_flipper_data <- penguins_clean %>% 
  filter_by_species("Adelie") %>% 
  subset_columns(c("body_mass_g", "flipper_length_mm", "species", "sex")) %>%  
  remove_NA()

# View filtered data
names(bodymass_flipper_data)
head(bodymass_flipper_data)

# Create exploratory plot for body mass and flipper length
exploratory_plot <- ggplot(data = bodymass_flipper_data, aes(x = body_mass_g, y = flipper_length_mm, color = sex)) +
  geom_point() +
    labs(title = "Body mass and flipper size for Adelie penguins", x = "Body mass (g)", y = "Flipper length (mm)", color = "Sex") +
    theme_bw() 

# Create a function to save plot as png file
agg_png("Figures/exploratory_plot.png", 
width = 20, height = 15, units = "cm", res = 600, scaling = 1)
exploratory_plot
dev.off()

print(exploratory_plot)
```

The exploratory plot above reveals two noteworthy trends:

- There appears to be a linear correlation between body mass and flipper length.

- The distributions of both variables appear to differ between the sexes.


### Hypotheses

1st hypothesis:
There is a significant positive correlation between the body mass and flipper length of Adelie penguins.

2nd hypothesis:
The difference between correlation coefficients for male and female penguins is significant.


### Statistical Methods

Linear regression model:

A linear regression analysis was used to investigate the relationship between body mass and flipper length. Body mass was treated as the independent variable, to quantify the extent to which variation in flipper length is explained by variation in body mass. The p-value produced by the linear model was compared to a significance level of 0.05.

Package used - Base R statistics


Step 2 - Fisher's Z-test

Fisher's Z-test was used to test whether the difference between correlation coefficients for each sex was significant. Pearson's correlation coefficient was calculated separately for males and females. The correlation coefficients were transformed into z-scores using the Fisher transformation. The transformed values approximately follow a normal distribution, making the comparison more appropriate for statistical testing. The test statistic was calculated from the z-scores, providing a standardized measure of the difference between correlation coefficients. Finally, the p-value was calculated and compared to a significance level of 0.05.


Packages used - Base R statistics, diffcor


```{r, echo=TRUE, warning=FALSE, message=FALSE}

# STEP 1 - Linear regression model 

# Run linear model 
linear_model <- lm(flipper_length_mm ~ body_mass_g, data = bodymass_flipper_data)

# Print summary 
summary(linear_model)



# STEP 2 - Fisher's Z-test 

# Install diffcor package
# install.packages("diffcor")

#Load diffcor package
library(diffcor)

# Partition data by sex using filter() function 
male <- filter(bodymass_flipper_data, sex == "MALE")
female <- filter(bodymass_flipper_data, sex == "FEMALE")

# Calculate correlation coefficients for each sex
cor_male <- cor(male$body_mass_g, male$flipper_length_mm)  
cor_female <- cor(female$body_mass_g, female$flipper_length_mm)  

# Calculate sample sizes
n_male <- nrow(male)  
n_female <- nrow(female)  

# Perform Fisher's Z-test for differences of correlations in two independent samples, using diffcor package
z_test_result <- diffcor.two(
  r1 = cor_male,
  r2 = cor_female,
  n1 = n_male,
  n2 = n_female,
  alpha = 0.05,
  cor.names = "Body Mass vs Flipper Length",
  alternative = "two.sided",
  digit = 3
)

# print result
print(z_test_result)

# r1 - male correlation coefficient
# r2 - female correlation coefficient
# z - test statistic for correlation difference in units of z distribution
# p - p value for two-sided test


```

### Results Figure

```{r, echo=TRUE, warning=FALSE, message=FALSE}

# Create results figure
results_figure <- ggplot(bodymass_flipper_data, aes(x = body_mass_g, y = flipper_length_mm, color = sex)) +
   geom_point() +
   geom_smooth(method = "lm", linewidth = 0.8, color = "black") +
  theme_bw() +
  labs(
    title = "Adelie Penguins: The Relationship between Body mass & Flipper Length",
    x = "Body Mass (g)",
    y = "Flipper Length (mm)",
    color = "Sex"
  ) +
  annotate(
    "text",
    x = Inf, y = -Inf, hjust = 2.6, vjust = -35,
    label = sprintf("Linear Model p-value: %s", format(summary(linear_model)$coefficients[2, 4], scientific = TRUE, digits = 2)),
  ) +
  annotate(
    "text",
    x = Inf, y = -Inf, hjust = 1.1, vjust = -7,
    label = sprintf("male coefficient: %.3f", cor_male),
  ) +
  annotate(
    "text",
    x = Inf, y = -Inf, hjust = 1.1, vjust = -5,
    label = sprintf("female coefficient: %.3f", cor_female),
  ) +
  annotate(
    "text",
    x = Inf, y = -Inf, hjust = 1.1, vjust = -3,
    label = sprintf("Z-test p-value: %s", z_test_result$p),
  ) +
  annotate(
    "text",
    x = Inf, y = -Inf, hjust = 2.4, vjust = -33,
    label = ("Positive correlation is significant")
  ) +
  annotate(
    "text",
    x = Inf, y = -Inf, hjust = 1.05, vjust = -1,
    label = ("Difference between sexes is not significant")
  )

print(results_figure)

# Create a function to save plot as png file
agg_png("Figures/results_figure.png", 
width = 20, height = 11, units = "cm", res = 600, scaling = 1)
results_figure
dev.off()


# Where the plot annotations appear (in Rmd / in the knitted file / in the saved image) seems to vary.
# The correct version of the graph is saved as a png in my GitHub Repository.
# The Graph above includes all the relevant statistical values.

```

### Results & Discussion

The results of the linear regression analysis revealed a strong positive relationship between body mass and flipper length for Adelie penguins (p < 0.05). This result therefore supports the 1st hypothesis.

An R-squared value of 0.2106 suggests that approximately 21.06% of variance in flipper length can be explained by changes in body mass. Therefore, although changes in body mass explain a significant proportion of the variance in flipper length, there are other factors contributing to this variation.

In addition to the linear regression analysis, Fisher's z-test was employed to assess whether the difference between correlation coefficients for males and females was significant. A p-value of 0.522 indicates that this is not the case, and so the 2nd hypothesis was rejected.

### Conclusion

To conclude, this study has provided convincing evidence to support the notion that body mass and flipper length are positively correlated in Adelie penguins. This relationship likely extends beyond the focal species of this study, and so wider conclusions about penguin physiology can be made with further research.

The sex-specific analysis of this relationship revealed that the strength of the correlation between these two variables is equal for males and females. This provides evidence to suggest that sexual dimorphisms do not impact the ratio of body mass to flipper length.


------------------------------------------------------------------------

## QUESTION 3: Open Science

### a) GitHub

https://github.com/anonbiologist/Reproducible_research.git

### b) Share your repo with a partner, download, and try to run their data pipeline.

https://github.com/biopenguinsox/Penguindata.git

### c) Reflect on your experience running their code. (300-500 words)

-   *What elements of your partner's code helped you to understand their data pipeline?*

The code is well laid out, with useful annotations at the end of each line describing in detail the action being carried out. The ggplot code was also partitioned with hash tags again describing what each component of the code was adding to the figure. The lines of code are also well spaced out to avoid confusion. 

The functions used to clean the raw data set were stored in a separate R file, in which they were well laid out and structured. By not defining the cleaning functions within the main script, it allowed for a concise pipe of cleaning functions that produces the same outcome but consumes far less space.

Finally, the structure of the ggplot code and the subsequent statistical test was easy to interpret, and the code could clearly be correlated with each component of the violin plots.

-   *Did it run? Did you need to fix anything?*

The only part of the code that did not run was the write.csv function, as the directory to which the data was saved was unique to that script. Although this line of code did give an error message, its purpose was to save the penguin data to a working directory, which is not essential but rather good practice. Therefore, I did not need to amend this but rather continued with the next line of code. Other than this, all the code ran without error, and all the plots were produced as intended.

-   *What suggestions would you make for improving their code to make it more understandable or reproducible, and why?*

My first suggestion is to make the instructions within the code more concise. Although they are informative, it does take up slightly too much space, especially in the ggplot code. Another minor point regarding the hash tags is that instructions are typically written before the line of code rather than after, although this does not impact the reproducibility of the study.

There are some instances where the code is quite dense and could benefit from more spacing, for example between the ggplot code and the png-saving code. This may improve how well one initially understands the code, as at first glance, they appear to be combined into one function.

Finally, both the exploratory and results figures could benefit from the inclusion of values. For the exploratory plot, the mean value for each species could be included. For the results plot, both the mean and confidence interval values could be included, so that the plot is not just a visual representation of the statistical analysis. It is also difficult to discern whether the confidence intervals of the Chinstrap and Gentoo penguins overlap, and so including the values for these bounds could improve understanding.

-   *If you needed to alter your partner's figure using their code, do you think that would be easy or difficult, and why?*

The data processing steps are clearly laid out, and use functions that are provided to the reader, which allows for flexibility of use. As for the ggplots, there are no elements that are unique to these plots, and so the code can be adapted without causing any issues to the plot layout. As no text or values were manually added to the figures, it provides simplicity when adapting the code to display the data differently.

### d) Reflect on your own code based on your experience with your partner's code and their review of yours. (300-500 words)

-   *What improvements did they suggest, and do you agree?*

My partner suggested that I included an explanation of all the functions used, and that in particular I did not explain all the components of my ggplot code. I agree that this hinders the reader from understanding my code, which could reduce the reproducibility of my study. To amend this, I would include hash tags within the ggplot code, to explain properly what each step is achieving.

It was also noted that the write.csv() function failed for my script, as it did when I attempted to run my partner's code. As this is not necessary for the reproducibility of the study but can improve understanding, I would amend this issue by placing a hash tag in front of this line of code, and perhaps signalling to the reader that it must be edited according to the working directory in use.

My partner suggested that I made the exploratory plot more visually interesting. I do not completely agree with this criticism, as I believe an exploratory plot is produced to indicate a trend and justify subsequent statistical analysis. I believe that my plot showed both a positive trend and a discrepancy between the sexes, which sufficiently led on to the statistical component of my analysis.

Finally, there appeared to be a problem with the layout of my results figure, as the labels did not consistently place where they were originally designed to be. Although I was able to save the correct version of the results figure to my GitHub repository, I agree that this inconsistency in formatting does reduce the reader's understanding as well as the reproducibility of the experiment. It was an oversight to assume that the figure would universally appear the same across formats. To amend this, I would use an alternate function to annotate() and not manually set the coordinates of my labels but rather fix them according to parameters that vary with the plot dimensions.

-   *What did you learn about writing code for other people?*

I learnt that in order to improve reproducibility, it is best to avoid personalizing code to your own script. By manually setting the coordinates of my figure annotations, I made it difficult for others to replicate my graph. I should have used parameters rather than hard-coded values. 

I have also learnt to consider whether my code is understandable to a new reader. There were instances where I assumed the reader would know a function I had used, and I now understand it is best to annotate and explain the elements of my ggplot code so that it can be reused without confusion.

Overall, I now understand that using clear and consistent variable names, compartmentalizing tasks with functions, and documenting actions with inline comments are all essential for reproducibility. In particular, I will now avoid manually formatting plots, so that my code is applicable beyond the original study for which it was written.



