

------------------------------------------------------------------------

## QUESTION 2: Data Pipeline

*Write a data analysis pipeline in your .rmd RMarkdown file. You should be aiming to write a clear explanation of the steps, the figures visible, as well as clear code.*

*Your code should include the steps practiced in the lab session:*

-   *Load the data*

-   *Appropriately clean the data*

-   *Create an Exploratory Figure (**not a boxplot**)*

-   *Save the figure*

-   ***New**: Run a statistical test*

-   ***New**: Create a Results Figure*

-   *Save the figure*

*An exploratory figure shows raw data, such as the distribution of the data. A results figure demonstrates the stats method chosen, and includes the results of the stats test.*

*Between your code, communicate clearly what you are doing and why.*

*Your text should include:*

-   *Introduction*

-   *Hypothesis*

-   *Stats Method*

-   *Results*

-   *Discussion*

-   *Conclusion*

*You will be marked on the following:*

### a) Your code for readability and functionality

### b) Your figures for communication

### c) Your text communication of your analysis

*Below is a template you can use.*

------------------------------------------------------------------------

### Introduction

The palmerpenguins dataset contains size measurements for three penguin species observed on three islands in the Palmer Archipelago, Antartica. Gentoo, Adelie, and Chinstrap penguins are the three extant members of the Pygoscelis genus, also known as Brush-tail penguins. Mitochondrial and nuclear DNA evidence suggest that the genus split from other penguins around 38 million years ago. Gentoo and Chinstrap penguins comprise a sister group, as Adelie penguins diverged from other members of the genus around 19 million years ago [3].

Penguins are flightless birds, with wings morphed by evolution into flippers, used to propel themselves underwater. It is reasonable to assume that larger penguins require larger flippers in order to create more powerful swimming strokes. This study aims to evidence this proposed association between body mass and flipper length.

Adelie penguins are a sexually dimorphic species, with adult males averaging larger than adult females in mass (~11%) and flipper length (~3%) [4]. As the male and female phenotype differs, it is possible that the strength of relationship between body mass and flipper length also differs between the sexes. An additional aim of this study is to test for a difference in correlation of these two variables, between male and female penguins.


References:

[3] Baker, A.J., Pereira, S.L., Haddrath, O.P. and Edge, K.A., 2006. Multiple gene evidence for expansion of extant penguins out of Antarctica due to global cooling. Proceedings of the Royal Society B: Biological Sciences, 273(1582), pp.11-17.

[4] Jennings, S., Varsani, A., Dugger, K.M., Ballard, G. and Ainley, D.G., 2016. Sex-based differences in Adélie penguin (Pygoscelis adeliae) chick growth rates and diet. PLoS One, 11(3), p.e0149090.

```{r Data Exploration}

# Install packages
install.packages(c("ggplot2", "palmerpenguins", "janitor", "dplyr", "ragg"))

# Load packages
library(ggplot2)
library(palmerpenguins)
library(janitor)
library(dplyr)
library(ragg)

# Load functions into rmd file
source("functions/cleaning.r")

# Export data from the palmerpenguins package
write.csv(penguins_raw, "data/penguins_raw.csv")

# Load data from working directory
penguins_raw <- read.csv("data/penguins_raw.csv")

# View raw data
names(penguins_raw)

# Clean data using pre-loaded cleaning functions
penguins_clean <- penguins_raw %>% 
  clean_column_names() %>%     # simplifies column names
  shorten_species() %>%        # simplifies species names
  remove_empty_columns_rows()  # Removes any empty columns or rows

# View cleaned data
names(penguins_clean)

# Save cleaned data to working directory
write.csv(penguins_clean, "data/penguins_clean.csv")

# Filter data using pre-loaded sub-setting functions
bodymass_flipper_data <- penguins_clean %>% 
  filter_by_species("Adelie") %>% 
  subset_columns(c("body_mass_g", "flipper_length_mm", "species", "sex")) %>%  
  remove_NA()

# View filtered data
names(bodymass_flipper_data)
head(bodymass_flipper_data)

# Create exploratory plot for body mass and flipper length
exploratory_plot <- ggplot(data = bodymass_flipper_data, aes(x = body_mass_g, y = flipper_length_mm, color = sex)) +
  geom_point() +
    labs(title = "Body mass and flipper size for Adelie penguins", x = "Body mass (g)", y = "Flipper length (mm)", color = "Sex") +
    theme_bw() 


# Create a function to save plot as png file
agg_png("Figures/exploratory_plot.png", 
width = 20, height = 15, units = "cm", res = 600, scaling = 1)
exploratory_plot
dev.off()


# Make sure your code prints. 
```

The exploratory plot above reveals two noteworthy trends:
1) There appears to be a linear correlation between body mass and flipper length.
2) The distributions of both variables appear to differ between the sexes.


### Hypotheses

1st:
There is a significant positive correlation between the body mass and flipper length of Adelie penguins.

2nd:
The difference between correlation coefficients for male and female penguins is significant.


### Statistical Methods

Linear regression model:

A linear regression analysis was used to investigate the relationship between body mass and flipper length. Body mass was treated as the independent variable, to quantify the extent to which variation in flipper length is explained by variation in body mass. The p-value produced by the linear model was compared to a signficance level of 0.05.

Package used - Base R statistics


Step 2 - Fisher's Z-test

Fisher's Z-test was used to test whether the difference between correlation coefficients for each sex was significant. Pearson's correlation coefficient was calculated separately for males and females. The correlation coefficients were transformed into z-scores using the Fisher transformation. The transformed values approximately follow a normal distribution, making the comparison more appropriate for statistical testing. The test statistic was calculated from the z-scores, providing a standardized measure of the difference between correlation coefficients. Finally, the p-value was calculated and compared to a significance level of 0.05.


Packages used - Base R statistics, diffcor


```{r Statistics}

# STEP 1 - Linear regression model 

# Run linear model 
linear_model <- lm(flipper_length_mm ~ body_mass_g, data = bodymass_flipper_data)

# Print summary 
summary(linear_model)



# STEP 2 - Fisher's Z-test 

# Install diffcor package
install.packages("diffcor")

#Load diffcor package
library(diffcor)

# Partition data by sex using filter() function 
male <- filter(bodymass_flipper_data, sex == "MALE")
female <- filter(bodymass_flipper_data, sex == "FEMALE")

# Calculate correlation coefficients for each sex
cor_male <- cor(male$body_mass_g, male$flipper_length_mm)  
cor_female <- cor(female$body_mass_g, female$flipper_length_mm)  

# Calculate sample sizes
n_male <- nrow(male)  
n_female <- nrow(female)  

# Perform Fisher's Z-test for differences of correlations in two independent samples, using diffcor package
z_test_result <- diffcor.two(
  r1 = cor_male,
  r2 = cor_female,
  n1 = n_male,
  n2 = n_female,
  alpha = 0.05,
  cor.names = "Body Mass vs Flipper Length",
  alternative = "two.sided",
  digit = 3
)

# print result
print(z_test_result)

# r1 - male correlation coefficient
# r2 - female correlation coefficient
# z - test statistic for correlation difference in units of z distribution
# p - p value for two-sided test


```

### Results Figure

```{r Plotting Results}

# Create results figure
results_figure <- ggplot(bodymass_flipper_data, aes(x = body_mass_g, y = flipper_length_mm, color = sex)) +
   geom_point() +
   geom_smooth(method = "lm", se = FALSE, linewidth = 0.8, color = "black") +
  theme_bw() +
  labs(
    title = "Adelie Penguins: The Relationship between Body mass & Flipper Length",
    x = "Body Mass (g)",
    y = "Flipper Length (mm)",
    color = "Sex"
  ) +
  annotate(
    "text",
    x = Inf, y = -Inf, hjust = 2.77, vjust = -30,
    label = sprintf("Linear Model p-value: %s", format(summary(linear_model)$coefficients[2, 4], scientific = TRUE, digits = 2)),
  ) +
  annotate(
    "text",
    x = Inf, y = -Inf, hjust = 1.15, vjust = -7,
    label = sprintf("ρ (♂): %.3f", cor_male),
  ) +
  annotate(
    "text",
    x = Inf, y = -Inf, hjust = 1.15, vjust = -5,
    label = sprintf("ρ (♀): %.3f", cor_female),
  ) +
  annotate(
    "text",
    x = Inf, y = -Inf, hjust = 1.1, vjust = -3,
    label = sprintf("Z-test p-value: %s", z_test_result$p),
  ) +
  annotate(
    "text",
    x = Inf, y = -Inf, hjust = 2.55, vjust = -28,
    label = ("Positive correlation is significant")
  ) +
  annotate(
    "text",
    x = Inf, y = -Inf, hjust = 1.04, vjust = -1,
    label = ("Difference between ρ(♂) & ρ(♀) is not significant")
  )


  
  
results_figure

# Create a function to save plot as png file
agg_png("Figures/results_figure.png", 
width = 20, height = 11, units = "cm", res = 600, scaling = 1)
results_figure
dev.off()

# Make sure your code prints. 

```

### Results & Discussion

The results of the linear regression analysis revealed a strong positive relationship between body mass and flipper length for Adelie penguins (p < 0.05). This result therefore supports the 1st hypothesis.

An R-squared value of 0.2106 suggests that approximately 21.06% of variance in flipper length can be explained by changes in body mass. Therefore, although changes in body mass explain a significant proportion of the variance in flipper length, there are other factors contributing to this variation.

In addition to the linear regression analysis, Fisher's z-test was employed to assess whether the difference between correlation coefficients for males and females was significant. A p-value of 0.522 indicates that this is not the case, and so the 2nd hypothesis was rejected.

### Conclusion

To conclude, this study has provided convincing evidence to support the notion that body mass and flipper length are positively correlated in Adelie penguins. This relationship likely extends beyond the focal species of this study, and so wider conclusions about penguin physiology can be made with further research.

The sex-specific analysis of this relationship revealed that the strength of the correlation between these two variables is equal for males and females. This provides evidence to suggest that sexual dimorphisms do not impact the ratio of body mass to flipper length.


------------------------------------------------------------------------


